---
title: "Pre-processing with phyloseq"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # send any figure output to this folder 
                      fig.path = "../../figures/Lecture/03_PreProcessing/") 
```

# Goals

The goal with this file is to pre-process any other contaminants and unexpected ASVs from our sample. We will need to: 

1. Remove Chloroplasts.
2. Remove Mitochondria. 
3. Removing ASVs from negative controls and also negative control samples.  
4. Evaluate accuracy by looking at the Mock Community.  
5. Check for reverse complements!
6. Remove samples without "enough" reads. 

## Input 

Here, we will start with the phyloseq object that we created in 02_AssignASVs.Rmd

1. **Raw phyloseq object**: We will need `data/02_AssignASVs/raw_phyloseq.RData`. 
2. **Mock Amplicon Fasta**: Please copy over the `mock_amplicons.fasta` from `/workdir/in_class_data/` to your `data/03_PreProcessing` folder.


## Output

1. We will write out a complex R data object known as an **S4** object, which in this case is known as a phyloseq output. We will store this in  `data/03_PreProcessing/preprocessed_phyloseq.RData`.  


# Set up the Environment 

## Set the seeds and threads 

```{r seed-threads}
# Set the seed for reproducibility
set.seed(238428)
```

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load packages  
```{r load-libraries}
# Load packages 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, install = FALSE)
```


# Load Phyloseq Data

```{r phyloseq-handoff}
# Load the data 
load("data/Lecture/02_AssignASVs_Lecture/raw_physeq.RData")

# Look at the data 
raw_physeq
```

# Clean up the data

Now, we can do some more cleaning! We know that there are organelles in eukaryotes derived from microbes, which still have a 16S that will contaminate our samples. Let's remove those! 

1. Chloroplasts
2. mitochondria  

```{r rm-mitos-chloros}
# Let's tak a quick look at the tax table 
# View(tax_df) # Let's see where chloroplasts are and also mitochondria

# Make new physeq without chloroplasts
noChloros_physeq <- 
  raw_physeq %>% 
  # rm chloroplasts, which are Order
  subset_taxa(Order != "Chloroplast" | is.na(Order))
  
# How many taxa were chloroplasts? 
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs

# Intuition chek 
#noChloros_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

# remove mitochondria 
noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASVs? 
num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs

# Let's take a quick look 
noChlorosMitos_physeq

# How many total asvs were removed from chloros and mitos? 
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)
# proportion of asvs kept? 
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

# Evaulate and remove the control samples 

Now, let's take a look at the negative controls and then make a decision about whether or not to remove the ASVs that we found in our controls 

1. Negative controls 
2. ASVs found within the negative controls and their distribution in the samples. 
3. Evaluate the mock community 

## Negative Controls 
```{r neg-controls}
# create a vector of samples that were negative controls 
control_samples <- c("WaterControl", "022um-Control", "3um-Control", "DNA-Ext-Control")

# Make a new phyloseq object of only the control samples 
control_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% control_samples) %>%
  # force remove ASVs with 0 counts 
  prune_taxa(taxa_sums(.) > 0, .)

# Let's take a look
control_physeq

# vector of control asvs
control_ASVs <- 
  control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()

# Visualize raw abundance values of control ASVs
control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

############# Check the control ASVs in the samples 
# Make new phyloseq object with only the control ASVs and ALL samples 
controlASV_physeq <- 
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% control_ASVs$ASV)

# Inspect it 
controlASV_physeq

# Let's take a look at the abundances 
controlASV_physeq %>%
  plot_bar(., "ASV", fill = "ASV", facet_grid = Sample_or_Control~.)

# Check tax of ASV_0003
#controlASV_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

# Would only like to keep ASV_0019 and ASV_0055, everything else is trash
controlASVs_toREMOVE <- dplyr::setdiff(control_ASVs$ASV, c("ASV_0019", "ASV_0055"))
# double check 
length(controlASVs_toREMOVE)

### Make my new phloseq object 
noControlsChlorosMitos_physeq <- 
  noChlorosMitos_physeq %>%
  # Actually remove the control ASVs 
  subset_taxa(., !(ASV %in% controlASVs_toREMOVE)) %>%
  # Remove control samples 
  subset_samples(., !(names %in% control_samples))

# Inspect no controls object 
noControlsChlorosMitos_physeq

# After filtering, how many ASVs were removed? 
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noControlsChlorosMitos_physeq)
total_ASVs_rm
```

# Evaluate Accuracy in the Mock Community 

- make a mock phyloseq object
- load in the standard fasta sequences of the mock community  
- test for differences between the two 


Before this next section, please make sure to copy over the `mock_amplicons.fasta` file from `/workdir/in_class_data/`

```{r}
# Make mock phyloseq object 
mock_physeq <- 
  noControlsChlorosMitos_physeq %>%
  subset_samples(., names == "MockZymoPos") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Inspect it 
mock_physeq
tax_table(mock_physeq)

# write us a messages summarizing this
cat("DADA2 inferred", ntaxa(mock_physeq), "ASVs present in the mock community.")

####### Load in the mock reference sequence from Zymo 
mock_ref <- getSequences("data/Lecture/03_PreProcessing_Lecture/mock_amplicons.fasta")
names(mock_ref)

### Pull out seqs from Sequenced Mock community
mock_seqs <- 
  mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseqs) 

# Inspect
head(mock_seqs)

# Which ASVs match the reference? 
matched_asvs <- 
  mock_seqs %>%
  rowwise() %>%
  # check each ASV if they match, TRUE, if not = FALSE
  mutate(Match = any(grepl(ASVseqs, mock_ref))) %>%
  # Create a vector of names that matched 
  pull(ASV)


# Evaluate which ones do NOT match?? 
# What's their abundances? 
# This help provide error rate 
cat(length(matched_asvs), "ASVs were exact matches to the expected reference sequence.")

# How many sequences were in the mock? 
mock_SeqDepth <- 
  mock_physeq %>%
  otu_table() %>%
  data.frame() %>%
  colSums()


# Who is in the sequenced mock community? 
mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(Genus, Species)
names(mock_ref)

# Curious: What are the mock counts in the actual samples 
noControlsChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% matched_asvs) %>%
  otu_table() %>%
  t()
```

What can we conclude about the mock community??? 
That all 8 ASVs in the mock exactly match 8 of 9 in the reference file. 

## Remove the mock community and it's ASVs

```{r rm-mock-asvs}
# Make a new phyloseq object without mock and its asvs
noControlsMock_ChlorosMitos_physeq <- 
  noControlsChlorosMitos_physeq %>%
  # remove the mock community 
  subset_samples(., names != "MockZymoPos") %>%
  # remove the ASVs from the mock 
  subset_taxa(., !(ASV %in% matched_asvs))

# Inspect
noControlsMock_ChlorosMitos_physeq

# Intuition Check 
num_ASVs_mock <- ntaxa(noControlsChlorosMitos_physeq) - ntaxa(noControlsMock_ChlorosMitos_physeq)

# Check 
stopifnot(num_ASVs_mock == 8)
```


# Reverse Complements 

```{r reverse-complements}
# Pull out Taxa Table
tax_table <- 
  noControlsChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 
```


# Evaluate the Sequencing Depth 

```{r seq-depth}
# The current data object
noControlsMock_ChlorosMitos_physeq

# Make a metadata_df for joining
metadata_df <- 
  noControlsMock_ChlorosMitos_physeq %>%
  sample_data() %>% 
  data.frame()

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  noControlsMock_ChlorosMitos_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_df, by = "names") 

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)
head(seqSums_df)

# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()

# plot it as a bar plot 
seqSums_df %>%
  ggplot(aes(x=reorder(names, TotalSeqs), y = TotalSeqs,
             fill = station)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample")

# histogram
seqSums_df %>%
  ggplot(aes(x= TotalSeqs, fill = station)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth")

# Density plot 
seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = station)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth")
```

# Remove samples with few reads 

```{r rm-samps}
# What's the min seq depth? 
min(sample_sums(noControlsMock_ChlorosMitos_physeq))

# Remove 20210615-MA-SCB2F
preprocessed_physeq <- 
  noControlsMock_ChlorosMitos_physeq %>%
  subset_samples(., names != "20210615-MA-SCB2F")

#What's the new min seq depth?
min(sample_sums(preprocessed_physeq))

# Final check of the sequencing depth for now
# Here, we could remove more samples if we needed to
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

# Save Preprocessed Phyloseq Object

## Add some metadata

We will add: 

1. A discrete column called `salinity_group` that is made from the continuous data as indicated below. 
2. The Sequencing Depth of each sample

We will add the salinity group based on the follow values. We will also make sure the metadata is in the correct order for analyses later.


| Salinity Group | Range (PSU) | Count |
|----------------|-------------|--------|
| Very Low       | 0–7         | 5      |
| Low            | >7–13       | 5      |
| Mid            | >13–20      | 11     |
| High           | >20         | 8      |


```{r add-metadata}
# Let's add a metadata column to the file for this dataset before moving on. 
# We will create a discrete, categorical variable called "salinity_group" out of the continuous form of the data. 
# 
metadata_df <- 
  preprocessed_physeq %>% 
  sample_data() %>%
  data.frame() %>%
  mutate(depth_cat = ifelse(depth == 0.0, "surface", "bottom"),
         # Create a new column for salinity group
         salinity_group = factor(
           cut(salinity_psu,
               breaks = c(-Inf, 7, 13, 20, Inf),
               labels = c("Very Low", "Low", "Mid", "High"),
               right = TRUE),
           # Make sure the factor order is correct
           levels = c("Very Low", "Low", "Mid", "High"), ordered = TRUE),
         # Fix the factor orders for date, station, and depth category 
         date = factor(date, levels = c("6/2/21", "6/15/21", "10/5/21"), ordered = TRUE),
         depth_cat = factor(depth_cat, levels = c("surface", "bottom"), ordered = TRUE),
         station = factor(station, levels = c("Copano West", "Copano East","Mesquite Bay",
                                              "Aransas Bay","Shipping Channel"), 
                          ordered = TRUE))

### Let's add the sequencing count data, so that doesn't get lost! 
# What are the sample size distributions?
num_seqs_df <- 
  preprocessed_physeq %>%
  sample_sums() %>%
  data.frame() %>%
  rownames_to_column(var = "names")

# Fix the column names 
colnames(num_seqs_df)[2] <- "Sequencing_Depth"

# Now merge it all 
metadata_df2 <- 
  metadata_df %>%
  left_join(num_seqs_df, by = "names")

## Check they match
stopifnot(metadata_df$names == metadata_df2$names)

# Rename the rows 
rownames(metadata_df2) <- metadata_df2$names

# Reassign to the phyloseq object, rewriting the old metadata 
sample_data(preprocessed_physeq) <- metadata_df2

# Double check
colnames(data.frame(sample_data(preprocessed_physeq)))

# Intuition check 
head(num_seqs_df)
```

## Finally, save the phyloseq object
```{r save-physeq}
# Please be sure to use the correct path to the appropriate directory in your file system
save(preprocessed_physeq, file = "data/Lecture/03_PreProcessing_Lecture/preprocessed_phyloseq.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```