---
title: "Pre-processing with phyloseq"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # send any figure output to this folder 
                      fig.path = "../figures/02_PreProcessing/") 
```

# Goals

The goal with this file is to pre-process any other contaminants and unexpected ASVs from our sample. We will need to: 

1. Remove Chloroplasts.
2. Remove Mitochondria. 
3. Removing ASVs from negative controls and also negative control samples.  
4. Evaluate accuracy by looking at the Mock Community.  
5. Check for reverse complements!
6. Remove samples without "enough" reads. 

## Input 

Here, we will start with the phyloseq object that we created in 02_Assign_ASVs.Rmd

1. **Metadata**: We will need `data/metadata.csv` and the file we created and saved from `02_AssignASVs.Rmd` and is located at `data/01_DADA2/track_read_counts.RData`. 
2. **ASV table**: We created and saved this file in `02_AssignASVs.Rmd` and it is located at `data/01_DADA2/ASV_table.tsv`. 
3. **Taxonomy Table**: We created and saved this file in `02_AssignASVs.Rmd` and it is located at `data/01_DADA2/ASV_taxonomy.tsv`. 


## Output

1. We will write out a complex R data object known as an **S4** object, which in this case is known as a phyloseq output. We will store this in  `data/02_PreProcessing/raw_preprocessed_phyloseq.RData`.  


# Set up the Environment 

## Set the seeds and threads 

```{r seed-threads}
# Set the seed for reproducibility
set.seed(238428)
```

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load packages  
```{r load-libraries}
# We will need BioStrings 
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("Biostrings")

# Now load everything else 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, install = FALSE)
```


# Load Data 

##  1. Metadata

Before starting this section, let's copy over the metadata 

- Where to copy to? `SalinityGradient_16S/data`
- Where to copy from?: `/workdir/in_class_data/SalinityGradient_16S/metadata.csv`


```{r load-track-reads}
# Load in the metadata
metadata_df <- read_csv("data/metadata.csv", col_names = TRUE) %>%
  # Add sample names to row names 
  rename("sample_names" = "...1") %>%
  # Let's also make a names column which will be used later to more easily access our sample names 
  mutate(names = sample_names)

# Inspect  
head(metadata_df)
dim(metadata_df)

# Then, load in track reads 
load("data/01_DADA2/track_read_counts.RData")

# Take a look at the data
head(track_counts_df)
dim(track_counts_df)

# Wait, the names are filenames! 
head(track_counts_df$sample_names)

# Fix the names! 
track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names, "_"), `[`,1) 

# Inspect to make sure it worked! 
head(track_counts_df$sample_names)

# Do they match?
length(track_counts_df$sample_names)
length(metadata_df$sample_names)

# It's not true! What's the difference? 
#setdiff(track_counts_df$sample_names, metadata_df$sample_names)

# Ohhh, this is an extra sample that accidentally got in. 
## Let's remove this sample by using left join, which will only merge with what we have in our metadata!
metadata_final_df <- 
  metadata_df %>%
  left_join(., track_counts_df, by = "sample_names") %>%
  # we need our sample names in rownames! 
  column_to_rownames(var = "sample_names")

# Intuition check 
dim(metadata_final_df)
head(metadata_final_df)
glimpse(metadata_final_df)
```



## 1. ASV Table
```{r load-asv-table}
# First, load asv table, df = dataframe
asv_df <- read.delim(file = "data/01_DADA2/ASV_table.tsv", sep = "\t",
                        # Add the column & row names
                        header = TRUE, row.names = 1) %>%
  # Remove that extra sample!
  select(-"CJ.V08.P")

# Inspect asv_df
head(asv_df)[,1:5]
dim(asv_df)

# Let's fix sample names, ^X removes an X at the beginning of the name 
colnames(asv_df) <- sub("^X", "", colnames(asv_df))
# And fix the "." to be a "-" 
  # Note that	"\\." → Matches a literal . (since . is a special character in regex)
colnames(asv_df)  <- gsub("\\.", "-", colnames(asv_df) )
  
# Inspect asv_df
head(colnames(asv_df)) # Now it looks good! 
head(asv_df)[,1:5]
```

##  Taxonomy Table
```{r load-tax-table}
tax_df <- read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t",
                     # Add the column & row names
                     header = TRUE, row.names = 1) 

# Inspect
head(tax_df)
dim(tax_df)

# Add a check to make sure ASV names match each other in the asv_df and the tax_df
stopifnot(rownames(asv_df) == rownames(tax_df))
```


# Handoff to phyloseq
```{r phyloseq-handoff}
# double check it's all good 
dim(asv_df)
dim(tax_df)

# The rownames need to between asv and tax tables
head(row.names(asv_df))
head(row.names(tax_df))
stopifnot(row.names(asv_df) == row.names(tax_df))

# The sample names need to match between metadata and asv tables
head(colnames(asv_df))
head(row.names(metadata_final_df))
stopifnot(colnames(asv_df) == row.names(metadata_final_df))

# Construct the phyloseq object 
raw_physeq <- phyloseq(otu_table(asv_df, taxa_are_rows = TRUE),
                       sample_data(metadata_final_df),
                       # The taxonomy table needs to first be converted into a matrix
                       tax_table(as.matrix(tax_df)))

# Check the phyloseq object!
raw_physeq

# Save this raw phyloseq object 
save(raw_physeq, file = "data/02_PreProcessing/raw_physeq.RData")
```

# Clean up the data

Now, we can do some more cleaning! We know that there are organelles in eukaryotes derived from microbes, which still have a 16S that will contaminate our samples. Let's remove those! 

1. Chloroplasts
2. mitochondria  

```{r rm-mitos-chloros}
# Let's tak a quick look at the tax table 
# View(tax_df) # Let's see where chloroplasts are and also mitochondria

# Make new physeq without chloroplasts
noChloros_physeq <- 
  raw_physeq %>% 
  # rm chloroplasts, which are Order
  subset_taxa(Order != "Chloroplast" | is.na(Order))
  
# How many taxa were chloroplasts? 
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs

# Intuition chek 
#noChloros_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

# remove mitochondria 
noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASVs? 
num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs

# Let's take a quick look 
noChlorosMitos_physeq

# How many total asvs were removed from chloros and mitos? 
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)
# proportion of asvs kept? 
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

# Evaulate and remove the control samples 

Now, let's take a look at the negative controls and then make a decision about whether or not to remove the ASVs that we found in our controls 

1. Negative controls 
2. ASVs found within the negative controls and their distribution in the samples. 
3. Evaluate the mock community 

## Negative Controls 
```{r neg-controls}
# create a vector of samples that were negative controls 
control_samples <- c("WaterControl", "022um-Control", "3um-Control", "DNA-Ext-Control")

# Make a new phyloseq object of only the control samples 
control_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% control_samples) %>%
  # force remove ASVs with 0 counts 
  prune_taxa(taxa_sums(.) > 0, .)

# Let's take a look
control_physeq

# vector of control asvs
control_ASVs <- 
  control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()

# Visualize raw abundance values of control ASVs
control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

############# Check the control ASVs in the samples 
# Make new phyloseq object with only the control ASVs and ALL samples 
controlASV_physeq <- 
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% control_ASVs$ASV)

# Inspect it 
controlASV_physeq

# Let's take a look at the abundances 
controlASV_physeq %>%
  plot_bar(., "ASV", fill = "ASV", facet_grid = Sample_or_Control~.)

# Check tax of ASV_0003
#controlASV_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

# Would only like to keep ASV_18 and ASV_56, everything else is trash
controlASVs_toREMOVE <- dplyr::setdiff(control_ASVs$ASV, c("ASV_0017", "ASV_0050"))
# double check 
length(controlASVs_toREMOVE)

### Make my new phloseq object 
noControlsChlorosMitos_physeq <- 
  noChlorosMitos_physeq %>%
  # Actually remove the control ASVs (except 18, 56)
  subset_taxa(., !(ASV %in% controlASVs_toREMOVE)) %>%
  # Remove control samples 
  subset_samples(., !(names %in% control_samples))

# Inspect no controls object 
noControlsChlorosMitos_physeq

# After filtering, how many ASVs were removed? 
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noControlsChlorosMitos_physeq)
total_ASVs_rm
```

# Evaluate Accuracy in the Mock Community 

- make a mock phyloseq object
- load in the standard fasta sequences of the mock community  
- test for differences between the two 


```{r}
# Make mock phyloseq object 
mock_physeq <- 
  noControlsChlorosMitos_physeq %>%
  subset_samples(., names == "MockZymoPos") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Inspect it 
mock_physeq
tax_table(mock_physeq)

# write us a messages summarizing this
cat("DADA2 inferred", ntaxa(mock_physeq), "ASVs present in the mock community.")

####### Load in the mock reference sequence from Zymo 
mock_ref <- getSequences("data/02_PreProcessing/mock_amplicons.fasta")
names(mock_ref)

### Pull out seqs from Sequenced Mock community
mock_seqs <- 
  mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseqs) 

# Inspect
head(mock_seqs)

# Which ASVs match the reference? 
matched_asvs <- 
  mock_seqs %>%
  rowwise() %>%
  # check each ASV if they match, TRUE, if not = FALSE
  mutate(Match = any(grepl(ASVseqs, mock_ref))) %>%
  # Create a vector of names that matched 
  pull(ASV)


# Evaluate which ones do NOT match?? 
# What's their abundances? 
# This help provide error rate 
cat(length(matched_asvs), "ASVs were exact matches to the expected reference sequence.")

# How many sequences were in the mock? 
mock_SeqDepth <- 
  mock_physeq %>%
  otu_table() %>%
  data.frame() %>%
  colSums()


# Who is in the sequenced mock community? 
mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(Genus, Species)
names(mock_ref)

# Curious: What are the mock counts in the actual samples 
noControlsChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% matched_asvs) %>%
  otu_table() %>%
  t()
```

What can we conclude about the mock community??? 
That all 8 ASVs in the mock exactly match 8 of 9 in the reference file. 

## Remove the mock community and it's ASVs

```{r rm-mock-asvs}
# Make a new phyloseq object without mock and its asvs
noControlsMock_ChlorosMitos_physeq <- 
  noControlsChlorosMitos_physeq %>%
  # remove the mock community 
  subset_samples(., names != "MockZymoPos") %>%
  # remove the ASVs from the mock 
  subset_taxa(., !(ASV %in% matched_asvs))

# Inspect
noControlsMock_ChlorosMitos_physeq

# Intuition Check 
num_ASVs_mock <- ntaxa(noControlsChlorosMitos_physeq) - ntaxa(noControlsMock_ChlorosMitos_physeq)

# Check 
stopifnot(num_ASVs_mock == 8)
```


# Reverse Complements 

```{r reverse-complements}
# Pull out Taxa Table
tax_table <- 
  noControlsChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 
```


# Evaluate the Sequencing Depth 

```{r seq-depth}
# The current data objec
noControlsMock_ChlorosMitos_physeq

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  noControlsMock_ChlorosMitos_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_final_df, by = "names") 

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)
head(seqSums_df)

# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()

# plot it as a bar plot 
seqSums_df %>%
  ggplot(aes(x=reorder(names, TotalSeqs), y = TotalSeqs,
             fill = station)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample")

# histogram
seqSums_df %>%
  ggplot(aes(x= TotalSeqs, fill = station)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth")

# Density plot 
seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = station)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth")
```

# Remove samples with few reads 

```{r rm-samps}
# What's the min seq depth? 
min(sample_sums(noControlsMock_ChlorosMitos_physeq))

# Remove 20210615-MA-SCB2F
raw_preprocessed_physeq <- 
  noControlsMock_ChlorosMitos_physeq %>%
  subset_samples(., names != "20210615-MA-SCB2F")

#What's the new min seq depth?
min(sample_sums(raw_preprocessed_physeq))

# Final check of the sequencing depth for now
# Here, we could remove more samples if we needed to
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

# Save Preprocessed Phyloseq Object
```{r save-physeq}
save(raw_preprocessed_physeq, file = "data/02_PreProcessing/raw_preprocessed_physeq.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```