---
title: "Preprocessing Data Analysis"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(fig.path = "../figures/03_PreProcessing")
```


# Goals 

1. Remove chloroplasts. 
2. Remove mitochondria 
3. Remove ASVs from negative controls. And, remove negative control samples. 
4. Evaluate sequencing accuracy with the mock community (positive control).
5. Look for reverse complements. 
6. Sequencing Depth: Remove samples without enough reads (obvious). 


# Input 

1. Raw phyloseq object. 

```
mkdir 02_AssignASVs # inside the data
cp /workdir/in_class_data/raw_physeq.RData /workdir/netID/Biodiversity_Salinity_16S/data/02_AssignASVs
```

# Output 

1. Cleaned up phyloseq object called `data/03_PreProcessing/preprocessed_physeq.RData`


# Set up the Environment

```{r load-packages-data}
set.seed(238428)

# Load packages 
pacman::p_load(phyloseq, tidyverse, install = FALSE)

# Load data 
load("data/02_AssignASVs/raw_physeq.RData")
# Take a look 
raw_physeq
```

# Clean up the data 

## Chloroplasts & Mitochondria 

```{r}
raw_physeq %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order))

# How to access data in a phyloseq object? 
#View(otu_table(raw_physeq))


