---
title: "Preprocessing Data Analysis"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(fig.path = "../figures/03_PreProcessing")
```


# Goals 

1. Remove chloroplasts. 
2. Remove mitochondria 
3. Remove ASVs from negative controls. And, remove negative control samples. 
4. Evaluate sequencing accuracy with the mock community (positive control).
5. Look for reverse complements. 
6. Sequencing Depth: Remove samples without enough reads (obvious). 


# Input 

1. Raw phyloseq object. 

```
mkdir 02_AssignASVs # inside the data
cp /workdir/in_class_data/raw_physeq.RData /workdir/netID/Biodiversity_Salinity_16S/data/02_AssignASVs
```

# Output 

1. Cleaned up phyloseq object called `data/03_PreProcessing/preprocessed_physeq.RData`


# Set up the Environment

```{r load-packages-data}
set.seed(238428)

# Load packages 
pacman::p_load(phyloseq, tidyverse, install = FALSE)

# Load data 
load("data/02_AssignASVs/raw_physeq.RData")
# Take a look 
raw_physeq
```

# Clean up the data 

## Chloroplasts & Mitochondria 

```{r}
noChlorosMitos_physeq <-   
  raw_physeq %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order)) %>% # 248 Chloroplast ASVs
  subset_taxa(Family != "Mitochondria" | is.na(Family)) # 74 Mito ASVs

# Take a look 
noChlorosMitos_physeq

# Intuition Check 
ntaxa(noChlorosMitos_physeq) / ntaxa(raw_physeq) 

# How to access data in a phyloseq object? 
#View(tax_table(raw_physeq))
```

# Negative Controls 

```{r neg-controls}
# create a vector of samples that were negative controls 
control_samples <- c("WaterControl", "022um-Control", "3um-Control", "DNA-Ext-Control")

# Create phyloseq object of ONLY negative controls 
control_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(names %in% control_samples) %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Visualize the ASV abundances in negative controls 
control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

# Vector of control ASVs (31 ASVs)
control_ASVs <- 
  control_physeq %>%
  taxa_names()

# Create a phyloseq object of just those 31 ASVs across 95 samples
controlASV_physeq <- 
  noChlorosMitos_physeq %>%
  subset_taxa(ASV %in% control_ASVs)

controlASV_physeq %>%
  plot_bar(., "ASV", facet_grid = Sample_or_Control~.)

# Check tax
#View(tax_table(controlASV_physeq))

# Remove 29 of the control ASVs from the data 
controlASvs_toRemove <- 
  dplyr::setdiff(control_ASVs, c("ASV_0019", "ASV_0055"))

length(controlASvs_toRemove)

# Make a new physeq object without these 29 ASVs 
noChlorosMitos_physeq %>%
  # actually rm the 29 ASVs
  subset_taxa(., !(ASV %in% controlASvs_toRemove)) %>%
  # remove the control samples 
  subset_samples(., !(names %in% control_samples))

```