---
title: "Biodiversity Analysis"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load Packages & Colors 

```{r load-packages-colors}
pacman::p_load(tidyverse, patchwork, iNEXT, phyloseq,
               install = FALSE)

# Load in colors for plotting 
source("code/colors_and_shapes.R")
```

# Load the Preprocessed Phyloseq Data

```{r load-data}
load("data/03_PreProcessing/preprocessed_phyloseq.RData")

# Take a look at the data 
preprocessed_physeq

# Intuition check on sequencing depth 
min(sample_sums(preprocessed_physeq))
```

# Calculate Biodiversity with Hill Numbers 

```{r calc-biodiv}
# Prepare input data 
## species = rows; samples = columns
iNEXT_input_df <- 
  preprocessed_physeq %>%
  otu_table() %>%
  data.frame()

# inspect
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:3]

# Run iNEXT 
#iNEXT_data <- 
#  iNEXT(iNEXT_input_df, q = c(0, 1, 2), datatype = "abundance")
load("data/04_Biodiversity/iNEXT_data.RData")

# Inspect Data 
str(iNEXT_data)
```

# Plot Rarefaction Curves

## Prepare Metadata
```{r prep-metadata}
# Make a metadata object
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()
# take a look
glimpse(metadata_df)

# Create Color dataframe for plotting 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
glimpse(color_df)

# Add a names column 
colnames(color_df)[1] <- "names"
head(color_df)

iNEXT_color_df <- 
  color_df %>%
  # fix sample names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = ""))  %>%
  # Merge with metadata 
  left_join(metadata_df, by = "names") %>%
  # Merge with colors 
  left_join(
    data.frame(salinity_colors = salinity_colors,
               salinity_group = names(salinity_colors)),
            by = "salinity_group")
# Intuition Check
head(iNEXT_color_df$salinity_colors)
```



```{r plot-rarefaction}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$salinity_colors) +
  scale_fill_manual(values = iNEXT_color_df$salinity_colors) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs",
       x = "Number of Sequences")
```

What conclusions can we draw from this plot? 

1. The y-axis values go down as we weight abundance from `q = 0, 1, 2`  
2. As abundance is weighted, the curve asymptotes faster. Meaning that richness is the *most sensitive* to sequencing depth whereas Simpsons